Algorithm explanation

### 2013-05-04 article/algo.tex
Introduced copy tree

The Horn clause solving algorithm receives an input problem in Horn
graph. A Horn graph is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two sets of vertices; Horn term vertices $V_T$
  and arrow vertices $V_\Rightarrow$. Each vertex $u \in
  V_\Rightarrow$ has eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, ..., x_{arity(P)})$ or a linear inequality $a_1
  x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}

@@

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u \xrightarrow{\theta} v$ over $V_T$ as follows.
\[ u \xrightarrow{\theta} v \Longleftrightarrow
\exists a \in V_\Rightarrow; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]

We may omit $\theta$ and write $u \rightarrow v$ for convenience for
later discussion. By the assumption that $G$ is acyclic,
$\rightarrow^+$ is a well-founded relation. We assume without loss of
generality that $V_T$ contains the greatest element $v_\bot$ with
respect to $\rightarrow^+$; otherwise we can always implicitly assume
$\varphi(v_\bot) = \top$ such that $\forall v \in V_T; v
\xrightarrow{\emptyset} v_\bot$.

@@

We slightly modify the definition $C_v$. Let $\hat V$ a subset of
vertices with multiple successors.

\begin{align*}
C_v = & \bigcup_{\substack{u \rightarrow v \\ u \not\in \hat V}} C_u \cup
\bigcup_{\substack{u \rightarrow v \\ u \in \hat V}} \textsc{Copy}_v (C_u)
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_{\substack{u \xrightarrow{\theta} v \\ u \not\in \hat V}} \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_{\substack{u \xrightarrow{\theta} v \\ u \in \hat V}}
  \mathbf{a}_{u,U \cup \left\lbrace v \right\rbrace, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_{\substack{u \rightarrow v \\ u \not\in \hat V}} b_{u,U} +
  \sum_{\substack{u \rightarrow v \\ u \in \hat V}} b_{u,U \cup \left\lbrace v \right\rbrace}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}

The procedure $\textsc{Copy}_v$ renames all free variables to avoid
name collision. Its definition is given as follows.

\[\textsc{Copy}_{\hat v}(\mathbf{a}_{u,V}) = \mathbf{a}_{u,(V \cup
  \left\lbrace \hat v \right\rbrace)} \]

All variables are then distinguished by the label of copy origin
nodes. This intuitively means that constraints generated for nodes in
$\hat V$ are duplicated at their successors. For consistency, the
variables without the label should be interpreted as that it has
$\emptyset$ for origin nodes.

When $C_{v_\bot}$ becomes satisfiable by specifing an appropriate
$\hat V$, a solution $\rho$ to $G$ is given as follows.

@@

If $C_{v_\bot}$ stays unsatisfiable, we can add a vertex with multiple
successors to $\hat V$ until all such nodes are in $\hat V$.

>>>>

The Horn clause solving algorithm receives an input problem in Horn
graph. A \textbf{Horn graph} is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two sets of vertices; Horn term vertices $V_T$
  and arrow vertices $V_\Rightarrow$. Each vertex $u \in
  V_\Rightarrow$ has eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, ..., x_{arity(P)})$ or a linear inequality $a_1
  x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}

@@

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u \xrightarrow{\theta} v$ over $V_T$ as follows.
\[ u \xrightarrow{\theta} v \Longleftrightarrow
\exists a \in V_\Rightarrow; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]
We may omit $\theta$ and write $u \rightarrow v$ for convenience for
later discussion. By the assumption that $G$ is acyclic,
$\rightarrow^+$ is a well-founded relation. We assume without loss of
generality that $V_T$ contains the greatest element $v_\bot$ with
respect to $\rightarrow^+$; otherwise we can always implicitly assume
$\varphi(v_\bot) = \top$ such that $\forall v \in V_T; v
\xrightarrow{\emptyset} v_\bot$.

@@

Let $\hat V_T$ a set of Horn term vertices $V_T$ with multiple
successors.
\[ \hat V_T = \left\lbrace v \in V_T \middle|
\exists u, u', \theta, \theta';
\left\lbrace (v,u,\theta), (v,u',\theta') \right\rbrace \subseteq E \wedge
(u \ne u' \vee \theta \ne \theta') \right\rbrace \]
We define \textbf{Copy tree}, which is a tree $T=(V_C,E_C,\xi)$
rooted at $v_{v_\bot}$.
\begin{itemize}
\item $V_C$ is a set of nodes.
\item $E_C$ is a set of labelled edges, which are expressed in the
  form $(u,v,e)$, where $e \in E \cup {\bot}$.
\item $\xi: V_C \rightarrow \hat V_T \cup \left\lbrace v_\bot
  \right\rbrace$ is a labelling function for nodes. The root node
  $v_{v_\bot}$ has a label $v_\bot$.
\end{itemize}

We provide $\textsc{Copy}_v$ procedure to rename all free variables
for avoiding name collision.
\[ \textsc{Copy}_v (x_V) = x_{V \cup \left\lbrace v \right\rbrace} \]
All variables are then distinguished by the label of copy origin
nodes. For consistency, the variables without the label should be
interpreted as that it has $\emptyset$ for origin nodes.

We give a constraint function over a copy tree
$\hat \Delta(v,e)$ for $v \in V_C, e \in E$ as
\[ \hat \Delta (v,e) = \textsc{Copy}_v (\Delta_{V'} (\xi (v)))
\text{ such that } V' = \left\lbrace v' \middle| (v, v',e) \in E_S \right\rbrace \]
extending $\Delta$ function to
$\Delta_V (v) = (\mathbf{a}_{v,\emptyset} \mathbf{x}_v \leq b_{v,\emptyset}, C_v)$
as
\begin{align*}
C_v = & \bigcup_\circ C_u \cup \bigcup_\bullet C_u \cup
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_\circ \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_\bullet \mathbf{a}_{u,U, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_\circ b_{u,U} +
  \sum_\bullet b_{u,U}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}
under
\[
\circ = \substack{
  u \rightarrow v \\
  \nexists u' \in V; \xi (u') = u \\
  \Delta_V (u) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)}
\qquad \bullet = \substack{
  u \rightarrow v \\
  \exists u' \in V; \xi (u') = u \\
  \hat \Delta (u', u \rightarrow v) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)} \]

This intuitively means that constraints are duplicated according to
the shape of copy tree $T$.
Given the constraint for the root node in the copy tree as
\[ \hat \Delta (v_{v_\bot}) = (\mathbf{a}_0 \mathbf{x}_{v_\bot} \leq b_0, C_{v_\bot}) \]
a model to $C_{v_\bot}$ gives a solution $\rho$ to $G$, if one exists.

@@

If $C_{v_\bot}$ stays unsatisfiable, we let $T$ grow based on the
unsatisfiable core.
